---
title: "Tutorial"
output: 
  learnr::tutorial:
    progressive: true
  
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

`AdhereR` is an R package that aims to facilitate the computing of adherence from EHD, as well as the transparent reporting of the chosen calculations. 
It contains a set of `R` `S3` *classes* and *functions* that *compute*, *summarize* and *plot* various estimates of adherence. 

This tutorial aims to introduce researchers to the principles of EHD preparation required to estimate adherence with the AdhereR package. It uses example data to illustrate the various decisions required and their impact on estimates, starting with the visualization of medication events, computation of persistence (treatment episode length), and computation of adherence.

Please contact us with suggestions, bug reports, comments (or even just to share your experiences using the package) either by e-mail (to Dan <ddediu@gmail.com>, Alexandra <alexadima@gmail.com> or Samuel (samuel.allemann@gmx.ch)) or using GitHub's reporting mechanism at our repository <https://github.com/ddediu/AdhereR>, which contains the full source code of the package. 

```{r quiz}
quiz(
  question("Which data can be used to estimate adherence with the AdhereR package?",
    answer("Pharmacy records", correct = TRUE),
    answer("MEMS data"),
    answer("Insurance claims", correct = TRUE),
    answer("Self-report")
  )
)
```

## Dataset

### Minimal requirements

`AdhereR` requires a dataset with a minimum of 3 variables for each medication event: *patient unique identifier*, *event date*, and *duration*. 
*Daily dosage* and *medication type* are optional. `AdhereR` is thus designed to use datasets that have already been extracted from EHD and prepared for calculation. 
The preliminary data preparation depend to a large extent on the specific database used and the type of medication and research design. 

Oftentimes, datasets can be large with hundreds of thousands or even millions of rows. Provided that working memory is big enough to hold the entire dataset, R can handle it. Manipulating these large datasets with the standard R libraries can be a hassle. We will use the package `data.table`, which usually requires less memory and is faster than the standard `data.frame` format. 

### Example dataset

Suppose you have a dataset with the following information:

- patient unique identifier (`PATIENT.ID`), 
- event date (`DATE.DISP`; date of dispensing, in the "mm.dd.yyyy" format),
- ATC code (`ATC.CODE`; alpha-numeric code to identify the medication)
- dosage (`DOSE`; strength of the dispensed medication with unit), and
- quantity (`QUANTITY`; number of units dispensed). 

Table 1 shows the medication events of one example patient: 32 medication events related to 8 medications. 
```{r, echo=TRUE, message=FALSE, warning=FALSE, results='asis'}
# Load the AdhereR and data.table libraries (and install if not already installed):
if (!require("AdhereR")) install.packages("AdhereR")
if (!require("data.table")) install.packages("data.table")

# Load example dataset
disp_events <- fread("data/example_disp_events.csv")
# Display the first patient as pretty markdown table:
knitr::kable(disp_events[PATIENT.ID == 42], caption = "<a name=\"Table-1\"></a>**Table 1.** Medication events for one example patient");
```

## Data preparation

### Data cleaning

First, we have to make sure that the data is in the right format. We can use the function `str()` to check the format of our variables and `summary()` for a first plausibility check.

```{r, echo=TRUE}
# Check format of variables:
str(disp_events)
```

We can see that the `DATE.DISP` column is in *CHARACTER* instead of *DATE* format, and the `DOSE` is in *CHARACTER* format, too, because the unit is appended to it. 

### Exercise: examine data

Write the command to display a summary of `disp_events`

```{r summary_events, exercise=TRUE}

```

### Conversion of data types

We want to convert the `DATE.DISP`to the appropriate format and extract the numeric part and the characters from the `DOSE`variable into separate variables `DOSE.num`and `UNIT`. 

```{r, echo=TRUE}
# Convert DATE to the DATE format and split DOSE into two variables:
disp_events[, DATE.DISP := as.Date(DATE.DISP, format = "%d.%m.%Y")] #convert Date to date format

disp_events[,c("DOSE.num", "UNIT"):= tstrsplit(DOSE, " ")] #split Dose on whitespace
disp_events[,DOSE := NULL]
```

### Exercise: type conversion

Write the commands to convert the `DOSE.num` variable to *NUMERIC* and the `UNIT` variable to *FACTOR*.

```{r convert_variables, exercise=TRUE, exercise.eval=TRUE}
# add the code to convert DOSE.num to numeric and UNIT to factor:



# Check format of variables:
str(disp_events)

# Check summary of variables:
summary(disp_events)
```

```{r convert_variables-hint}
disp_events[,`:=` (DOSE.num = as.numeric(DOSE.num) #convert DOSE.num to numeric
                   )]
```

Now that the data is in the right format, we can see if there are any implausible or missing data from the summary.
We can see that all the dates appear to be from the years 2014-2016, which corresponds with our intended follow-up window. There are 31 different medications (as seen from the 31 factor levels of the `ATC.CODE`variable).

### Reproducible and transparent data modifications

There are 4 different units in the `UNIT` variable: *MG*, *MCG*, *MICROG*, and *UI*. *MCG* and *MICROG* both refer to the same unit, microgrammes, so there should be only one version of this unit. We could change the original data and replace all instances of *MCG* with *MICROG*, but one of the data cleaning principles is to never change the original data. Instead, we will modify the data transparently and reproducibly in our script. If there are a lot of modificiations, this could be in a separate file that contains only the modifications.

```{r, echo=TRUE}
# Assign *MCG* and *MICROG* to the same factor level
levels(disp_events$UNIT) <- list(MICROG=c("MCG", "MICROG"), MG="MG", UI="UI")
```

## Data processing

### Supply duration

To calculate CMAs, `AdhereR` requires a `DURATION` for each dispensing event, but we only have the quantity. We could assume that patients need to administer one unit per day and use the `QUANTITY` variable, but oftentimes, this is not appropriate. Sometimes, standard doses, e.g. WHO's 'Defined Daily Dose' or other assumptions may be appropriate in some instances, but might introduce bias in other situations. For this example, we can use a second database where the prescribed dosage for each medication per patient is recorded: 

- patient unique identifier (`PATIENT.ID`), 
- event date (`DATE.PRESC`; date of prescription, in the "mm/dd/yyyy" format),
- ATC code (`ATC.CODE`; alpha-numeric code to identify the medication)
- dosage (`DAILY.DOSE`; prescribed dose of the medication per day), and
- unit (`UNIT`; unit of the prescribed dose). 

```{r, echo=TRUE}
# Load example prescription data:

load("data/example_presc_events.RData")

presc_events

```
Conveniently, this dataset is already clean and all the data is in the right format. Moreover, there is only one prescription event per medication, which occured before the first dispensing event, so we don't have to deal with prescription changes during our follow-up period.

### Exercise: Add daily doses from separate file

We can now merge the two datasets and calculate the duration for each dispensing event. We merge by `PATIENT.ID`, `ATC.CODE` code and `UNIT` to make sure that events are matched correctly. This is why it was necessary to clean up the units: Otherwise, some events might not merge correctly due to mismatches between the units.

By default, the `merge` function only includes rows where the ID-variable are present in both instances. This means that we only capture medications that were prescribed and at least once dispensed during the follow-up period. If we want to capture all events, we can specify `all = TRUE` in the function arguments.

Write the command to merge `disp_events`and `presc_events` by `PATIENT.ID`, `ATC.CODE` code and `UNIT` into a new object called `med_events`

```{r merge-data, exercise=TRUE, exercise.eval=TRUE}
# Merge dispensing and prescription data to create an object 'med_events':

# Display the merged data.table 'med_events'

```

```{r merge-data-solution}
med_events <- merge(disp_events, presc_events, by = c("PATIENT.ID", "ATC.CODE", "UNIT"), all = TRUE, sort = FALSE)

med_events
```

### Exercise: Calculate supply duration

Now calculate the supply duration by multiplying the dispensed Dose and quantity and dividing it by the prescribed daily dose.

```{r, echo = FALSE}
disp_events[,`:=` (DOSE.num = as.numeric(DOSE.num), #convert DOSE.num to numeric
                   UNIT = as.factor(UNIT) #convert UNIT to factor variable
                   )]

med_events <- merge(disp_events, presc_events, by = c("PATIENT.ID", "ATC.CODE", "UNIT"), all = TRUE, sort = FALSE)
```


```{r calculate-duration, exercise = TRUE, exercise.eval=TRUE}
# Add a new column DURATION and Calculate the supply duration:

# Display a summary of the object 'med_events'

```
```{r calculate-duration-solution}
med_events[,DURATION := (DOSE.num*QUANTITY)/DAILY.DOSE]

summary(med_events)
```

### Multiple data sources

In its newest version, `AdhereR` now offers a function to link dispensing, prescription, and hospitalization data to improve the accuracy of adherence estimation. For each dispensing event, it:

1. automatically selects the last prescibed dose to calculate supply duration,
2. checks for prescription changes and hospitalizations during this period, and
3. adjusts the duration accordingly.

It requires the following input:

- `x` : A *data.frame* with the dispensing data
- `y` : A *data.frame* with the prescription data
- `z` : optional, a *data.frame* with the hospitalization data
- `ID.var` : A *character vector* of the ID column (identical in all data sources)
- `DATE.PRESC.var` : A *character vector* of the prescription date column (in y)
- `DATE.DISP.var` : A *character vector* of the prescription date column (in x)
- `DATE.format` : A *character vector* of the date format (identical in all data sources)
- `CATEGORY.var` : A *character vector* of the medication identification column (identical for x and y)
- `TOTAL.DOSE.var` : A *numeric vector* of the column with the dispensed dose (in x)
- `PERDAY.var` : A *numeric vector* of the column with the daily prescribed dose (in y)
- `PRESC.DURATION.var` : : optional, A *integer vector* of the column with the prescription duration in days (in y)
- `UNIT.var` : optional, A *character vector* of the medication unit column (identical for x and y)
- `FORM.var` : optional, A *character vector* of the medication form column (identical for x and y)
- `VISIT.var` : optional, A *integer vector* of the visit number (in y)
- `force.init.presc` : *logical*, default *TRUE*; should first prescibed dose be used for dispensing events occuring before the first prescription event?
- `force.presc.renew` : *logical*, default *TRUE*; if a medication has not been prescribed during a prescription event, should its prescription end on this date?
- `consider.dosage.change` : *logical*, default *TRUE*; should the supply duration be recalculated in case of prescription changes?

### Exercise: link dispensing and prescription data with AdhereR

Use the `medication_match` function to merge `disp_events` and `presc_events` into `med_events`.

```{r link-multiple-sources, exercise=TRUE}
# Merge dispensing and prescription data with AdhereR's medication_match function:

```

In addition to the standard AdhereR columns, the output of the matching function contains some more columns with additional information:

- `FIRST.PRESC` : A *Date* column with the date when the treatment was first prescribed 
- `PRESC.START` : A *Date* column with the start date of a prescription episode
- `PRESC.END` : A *Date* column with the end date of a prescription episode. If there is no end date, this will be `NA`
- `DOSAGE.CHANGE` : An *integer* column with the number of dosage changes considered for a given dispensing event.

## Visualization of patient records

### Explore medication histories 

A first step towards deciding which algorithm is appropriate for these data is to explore medication histories visually. 

We can do this by creating an object of type `CMA0` for the two example patients, and plotting it. 
This type of plots can of course be created for a much bigger subsample of patients and saved as as a `JPEG`, `PNG`, `TIFF`, `EPS` or `PDF` file using `R`'s plotting system for data exploration.

```{r, echo=TRUE}
# Create an object "cma0" of the most basic CMA type, "CMA0":
# example_data <- as.data.frame(na.omit(med_events[PATIENT.ID %in% c(42, 43), ]))
# 
# cma0 <- CMA0(data=example_data, # use the two selected patients
#              ID.colname="PATIENT.ID", # the name of the column containing the IDs
#              event.date.colname="DATE.DISP", # the name of the column containing the event date
#              event.duration.colname="DURATION", # the name of the column containing the duration
#              event.daily.dose.colname="DAILY.DOSE", # the name of the column containing the dosage
#              medication.class.colname="ATC.CODE", # the name of the column containing the category
#              followup.window.start=0,  # FUW start in days since earliest event
#              observation.window.start=182, # OW start in days since earliest event
#              observation.window.duration=365 # OW duration in days
#              date.format="%Y-%m-%d"
# ); # date format (mm/dd/yyyy)

example_data <- med.events[med.events$PATIENT_ID %in% c(37, 76), ]

cma0 <- CMA0(data=example_data, # use the two selected patients
             ID.colname="PATIENT_ID", # the name of the column containing the IDs
             event.date.colname="DATE", # the name of the column containing the event date
             event.duration.colname="DURATION", # the name of the column containing the duration
             event.daily.dose.colname="PERDAY", # the name of the column containing the dosage
             medication.class.colname="CATEGORY", # the name of the column containing the category
             followup.window.start=0,  # FUW start in days since earliest event
             observation.window.start=182, # OW start in days since earliest event
             observation.window.duration=365, # OW duration in days
             date.format="%m/%d/%Y"); # date format (mm/dd/yyyy)

# Plot the object (CMA0 shows the actual event data only):
plot(cma0, # the object to plot
     align.all.patients=TRUE
     )
```

### Interactive plotting

```{r}
plot_interactive_cma(data=example_data,
                     cma.class="per episode",
                     ID.colname="PATIENT_ID",
                     event.date.colname="DATE",
                     event.duration.colname="DURATION",
                     event.daily.dose.colname="PERDAY",
                     medication.class.colname="CATEGORY",
                     date.format="%m/%d/%Y")
```

## Adherence estimation

### Adherence, Initiation, Implementation, and Persistence

`AdhereR` can facilitate all aspects of adherence. For a better flow of this tutorial, we will look at each element in the following order:

1. Initiation
2. Adherence
3. Persistence
4. Implementation

##  Initiation

If we have a dispensing and a prescription database, we can calculate time to initiation from the output of the `medication_match` function. 

## Adherence

`AdhereR`estimates adherence as *Continuous Medication Availability* (CMA). The *simple* CMA measures *CMA1* - *CMA9* do not differentiate between implementation and persistence

If we cannot differentiate between persistence and implementation, we can still calculate an overall adherence value

### Exercise: Calculate adherence

Display the medication events and adherence estimates for the patient with the ID 42. The follow-up window starts 30 days after the first event and the observation window covers 120 days until 300 days after the first event.

```{r visualize-events, exercise = TRUE}

```

## Persistence

To distinguish between persistence with treatment and quality of implementation, `AdhereR` provides the function `compute.treatment.episodes()`. It reports the patient ID, episode number, date of episode start, gap days at the end of or after the treatment episode, duration of episode, and episode end date.

### Exercise: compute treatment episodes

Use the `compute.treatment.episodes()` function to compute treatment episodes from the first event until the end of the follow-up period (720 days) with the following conditions:

- Change of medication type should be considered as a new treatment episode
- A gap of 90 days or more should be considered as a treatment interruption/discontinuation
- Existing supply will be finished before starting a new supply of the same medication, but not for different medications
- If the dosage recommendation changes with a new supply, existing supply will be administered according to the new dosage recommendations
```{r treatment-episodes, exercise = TRUE, exercise.eval=TRUE}

```

## Implementation

### Calculation for multiple treatment episodes during an observation window

```{r treatment-episodes-solution}
TEs <- compute.treatment.episodes(example_data,
                                  ID.colname="PATIENT_ID",
                                  event.date.colname="DATE",
                                  event.duration.colname="DURATION",
                                  event.daily.dose.colname="PERDAY",
                                  medication.class.colname="CATEGORY",
                                  carryover.within.obs.window = TRUE, # carry-over inside the OW
                                  carry.only.for.same.medication = TRUE, # & only for same type
                                  consider.dosage.change = TRUE, # dosage change starts new episode...
                                  medication.change.means.new.treatment.episode = TRUE, # & type change
                                  maximum.permissible.gap = 90, # & a gap longer than 180 days
                                  maximum.permissible.gap.unit = "days", # unit for the above (days)
                                  followup.window.start = 0, # 2-years FUW starts at earliest event
                                  followup.window.start.unit = "days",
                                  followup.window.duration = 720,
                                  followup.window.duration.unit = "days",
                                  date.format = "%m/%d/%Y")

TEs
```


```{r}
cmaE <- CMA_per_episode(CMA="CMA7", # apply the simple CMA7 to each treatment episode
                        data=example_data,
                        ID.colname="PATIENT_ID",
                        event.date.colname="DATE",
                        event.duration.colname="DURATION",
                        event.daily.dose.colname="PERDAY",
                        medication.class.colname="CATEGORY",
                        carryover.within.obs.window = TRUE,
                        carry.only.for.same.medication = FALSE,
                        consider.dosage.change = FALSE, # conditions on treatment episodes
                        medication.change.means.new.treatment.episode = TRUE,
                        maximum.permissible.gap = 180,
                        maximum.permissible.gap.unit = "days",
                        followup.window.start=0,
                        followup.window.start.unit = "days",
                        followup.window.duration = 365 * 2,
                        followup.window.duration.unit = "days",
                        observation.window.start=0,
                        observation.window.start.unit = "days",
                        observation.window.duration=365*2,
                        observation.window.duration.unit = "days",
                        date.format="%m/%d/%Y")

plot(cmaE, patients.to.plot=c("76"), show.legend=FALSE)
```




