---
title: "Tutorial for AdhereR v0.5"
output: 
  learnr::tutorial:
    progressive: true
  
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

library(AdhereR)
library(data.table)

durcomp.dispensing <- AdhereR::durcomp.dispensing
durcomp.prescribing <- AdhereR::durcomp.prescribing

load("data/event_durations.rda")

```

## Introduction

`AdhereR` is an R package that aims to facilitate the computing of adherence from EHD, as well as the transparent reporting of the chosen calculations. 
It contains a set of `R` `S3` *classes* and *functions* that *compute*, *summarize* and *plot* various estimates of adherence. 

This tutorial aims to introduce researchers to the principles of EHD preparation required to estimate adherence with the AdhereR package. It uses example data to illustrate the various decisions required and their impact on estimates, starting with the visualization of medication events, computation of persistence (treatment episode length), and computation of adherence.

Please contact us with suggestions, bug reports, comments (or even just to share your experiences using the package) either by e-mail (to Dan <ddediu@gmail.com>, Alexandra <alexadima@gmail.com> or Samuel <samuel.allemann@gmx.ch>) or using GitHub's reporting mechanism at our repository <https://github.com/ddediu/AdhereR>, which contains the full source code of the package. 

```{r quiz-1}
quiz(
  question("Which data can be used to estimate adherence with the AdhereR package?",
    answer("Prescription data", correct = TRUE),
    answer("Pharmacy records", correct = TRUE),
    answer("MEMS data"),
    answer("Insurance claims", correct = TRUE),
    answer("Self-report"),
    incorrect = "At the moment, AdhereR only works with administrative data, such as prescription or dispensing records and insurance claims."
  )
)
```

## Dataset

### Minimal requirements

`AdhereR` requires a clean dataset with a minimum of information for each medication event. `AdhereR` is thus designed to use datasets that have already been extracted from EHD and prepared for calculation. 
The preliminary data preparation depend to a large extent on the specific database used and the type of medication and research design. 

```{r quiz-2}
quiz(
  question("Which minimum information is required to estimate adherence with AdhereR?",
    answer("Unique patient ID", correct = TRUE),
    answer("Event date", correct = TRUE),
    answer("Daily dosage"),
    answer("Medication type"),
    answer("Supply duration", correct = TRUE),
    incorrect = "Daily dosage and Medication type is optional and only required for some algorithms."
  )
)
```

Oftentimes, datasets can be large with hundreds of thousands or even millions of rows. Provided that working memory is big enough to hold the entire dataset, R can handle it. `AdhereR` can even access data (read and write) stored in various types of databases, ranging from "classic" relational databases (such as `MySQL` and `SQLite`) to new approaches (such as Apache's `Hadoop`). This allows the seamless processing of very large datasets across many types of architectures, ranging from a consumer-grade laptop to large heterogeneous computer clusers, without loading the dataset in memory.

When working in memory, manipulating large datasets with the standard R libraries can be a hassle. We propose to use the package `data.table`, which usually requires less memory and is faster than the standard `data.frame` format. 

### Example dataset

Suppose you have a dataset with the following information:

- patient unique identifier (`ID`), 
- event date (`DATE.DISP`; date of dispensing, in the "yyyy-mm-dd" format),
- ATC code (`ATC.CODE`; alpha-numeric code to identify the medication),
- Unit (`UNIT`; the unit of the dosage strength),
- Form (`FORM`; the galenic form of the preparation),
- dosage (`DOSE`; strength of the dispensed medication with unit), and
- quantity (`TOTAL.DOSE`; total dose (in units of `UNIT`) dispensed). 

`AdhereR` contains an example dataset with just this information: `durcomp.dispensing`

Table 1 shows the medication events of one example patient in `durcomp.dispensing`: 32 medication events related to 8 medications. 
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Load the AdhereR and data.table libraries (and install if not already installed):
if (!require("AdhereR")) install.packages("AdhereR")
if (!require("data.table")) install.packages("data.table")

# Display the data for the first patient:
durcomp.dispensing[ID == 1]
```

## Data preparation

### Data cleaning

First, we have to make sure that the data is in the right format. We can use the function `str()` to check the format of our variables and `summary()` for a first plausibility check.

```{r, echo=TRUE}
# Check format of variables:
str(durcomp.dispensing)
```

`R` treats the columns in a data.table as objects with a specific `class` attribute. The `class` describes the format of the column, similar to the *Cell Format* in Excel spreadsheets. In our dataset, there are two *numeric* columns: `ID` and `TOTAL.DOSE`. `DATE.DISP` is in *Date* format. And `ATC.CODE`, `UNIT`, and `FORM` columns are in *CHARACTER* format. 

### Exercise: examine data

Write the command to display a summary of `durcomp.dispensing`. Run your code by clicking on "Run Code" at the top right of the Code window. If you want to start over, click "Start Over", at the top left. Click the "Hint" button for Help.

```{r summary_events, exercise=TRUE}
# Display summary of data:

```

<div id="summary_events-hint">
**Hint:** You may want tu use the `summary()` function.
</div>

### Conversion of data types

Another special `R` class ist *factor*, which can be used for categorical data. It groups the data into a limited set of different *levels*, for example *male* and *female*.
We want to convert the `ATC.CODE`, `UNIT`, and `FORM` columns to *FACTOR*. We can directly manipulate data in a `data.table`, using the following syntax: `DT[i, j, by]`. This tells `R` to take `DT` (the `data.table`), subset/reorder rows using `i`, then do `j`, grouped by `by`. In `i`, `j`, and `by`, we can referr to any column within `DT` using its name. We don't have to use all of `i`, `j`, and `by`; if we want to do something with all rows without grouping, we can just use `j` without `i` and `by`. An excellent introduction to `data.table` can be found [HERE](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html){target="_blank"}.

To convert a column to a new class, we can do it directly inside the `data.table`, using the operator `:=`:
```{r, echo=TRUE}
# Convert DATE to the DATE format and split DOSE into two variables:
durcomp.dispensing[, ATC.CODE := as.factor(ATC.CODE)] #convert ATC.CODE to factor
```

### Exercise: type conversion

Write the commands to convert the `UNIT`, and `FORM` columns to *FACTOR*.

```{r convert_variables, exercise=TRUE, exercise.eval=TRUE}
# add the code to convert UNIT and FORM to factor:


```

```{r convert_variables-hint}
# To manipulat multiple columns at the same time, use the following format:
# DT[,`:=` (col1 = do(x), 
#           col2 = do(y))]
```

Now that the data is in the right format, we can examine them for any implausible or missing data, using `summary`.
For example, we can see that there are three different units: *MG*, *MICROG*, and *UI*. 

## Data processing

### Event durations

To calculate CMAs, `AdhereR` requires a `DURATION` for each medication event, but we only have the total dispensed dose. If medications are not supplied for fixed durations but as a quantity that may last for various durations based on the prescribed dose, the supply duration has to be calculated based on dispensed and prescribed doses. Standard doses, e.g. WHO's *Defined Daily Dose* or other assumptions may be appropriate in some instances, but might introduce bias in other situations.

With AdhereR comes a second sample database, `durcomp.prescribing`, with prescriptions for the same patients as in `durcomp.dispensing`: 

- patient unique identifier (`ID`), 
- event date (`DATE.PRESC`; date of prescription, in the "mm/dd/yyyy" format),
- visit number (`VISIT`; the consecutive number of the prescription instances),
- ATC code (`ATC.CODE`; alpha-numeric code to identify the medication),
- Unit (`UNIT`; the unit of the dosage strength),
- Form (`FORM`; the galenic form of the preparation),
- prescription duration (`PRESC.DURATION` ; the duration (in days) for which the prescription is intended. Can be `NA` if the prescription is continuous without a fixed end date.), and
- dosage (`DAILY.DOSE`; prescribed dose of the medication per day). 

Table 2 shows the prescriptions for the first patient in `durcomp.prescribing`: 
```{r, echo=FALSE}
# Load example prescription data:

durcomp.prescribing[ID==1]

```

<!-- ### Merge prescription and dispensing data -->

<!-- We can now merge the two datasets and calculate the duration for each dispensing event. We merge by `PATIENT.ID`, `ATC.CODE` code and `UNIT` to make sure that events are matched correctly. This is why it was necessary to clean up the units: Otherwise, some events might not merge correctly due to mismatches between the units. -->

<!-- By default, the `merge` function only includes rows where the ID-variable are present in both instances. This means that we only capture medications that were prescribed and at least once dispensed during the follow-up period. If we want to capture all events, we can specify `all = TRUE` in the function arguments. -->

<!-- ### Exercise: Add daily doses from separate file -->

<!-- Write the command to merge `disp_events` and `presc_events` by `PATIENT.ID`, `ATC.CODE` code and `UNIT` into a new object called `med_events` -->

<!-- ```{r merge-data, exercise=TRUE, exercise.eval=TRUE} -->
<!-- # Merge dispensing and prescription data to create an object 'med_events': -->

<!-- # Print the merged data.table 'med_events' -->

<!-- ``` -->

<!-- ```{r merge-data-solution} -->
<!-- # Merge dispensing and prescription data to create an object 'med_events': -->
<!-- med_events <- merge(disp_events, presc_events, by = c("PATIENT.ID", "ATC.CODE", "UNIT"), all = TRUE, sort = FALSE) -->

<!-- # Print the merged data.table 'med_events' -->
<!-- med_events -->
<!-- ``` -->

<!-- ### Exercise: Calculate supply duration -->

<!-- Now calculate the supply duration by multiplying the dispensed Dose and quantity and dividing it by the prescribed daily dose. -->

<!-- ```{r, echo = FALSE} -->
<!-- disp_events[,`:=` (DOSE.num = as.numeric(DOSE.num), #convert DOSE.num to numeric -->
<!--                    UNIT = as.factor(UNIT) #convert UNIT to factor variable -->
<!--                    )] -->

<!-- med_events <- merge(disp_events, presc_events, by = c("PATIENT.ID", "ATC.CODE", "UNIT"), all = TRUE, sort = FALSE) -->
<!-- ``` -->


<!-- ```{r calculate-duration, exercise = TRUE, exercise.eval=TRUE} -->
<!-- # Add a new column DURATION and Calculate the supply duration: -->

<!-- # Display a summary of the object 'med_events' -->

<!-- ``` -->
<!-- ```{r calculate-duration-solution} -->
<!-- # Add a new column DURATION and Calculate the supply duration: -->
<!-- med_events[,DURATION := (DOSE.num*QUANTITY)/DAILY.DOSE] -->

<!-- # Display a summary of the object 'med_events' -->
<!-- summary(med_events) -->
<!-- ``` -->

### Multiple data sources

`AdhereR` offers functions to link dispensing, prescription, and other data to improve the accuracy of adherence estimation. 
The main function `compute_event_durations` performs iterations of the following steps for each dispensing event:

1. automatically selects the last prescibed dose to calculate supply duration,
2. checks for prescription changes, treatment interruptions, and special periods such as hospitalizations during the calculated supply duration, and
3. adjusts the duration according to arguments provided by the user.

The function requires the following input:

- `disp.data` : A *data.frame* with the dispensing data
- `presc.data` : A *data.frame* with the prescription data
- `special.periods.data` : optional, a *data.frame* with special periods data
- `ID.colname` : The quoted name of the ID column (identical in all data sources)
- `medication.class.colnames` : A *character vector* with the name(s) of the medication identification column(s) (identical for `disp.data` and `presc.data`)
- `disp.date.colname` : The quoted name of the dispensing date column (in `disp.data`)
- `total.dose.colname` : The quoted name of the column with the dispensed dose (in `disp.data`)
- `presc.date.colname` : The quoted name of the prescription date column (in `presc.data`)
- `presc.daily.dose.colname` : The quoted name of the column with the daily prescribed dose (in `presc.data`)
- `presc.duration.colname` : : The quoted name of the the column with the prescription duration in days (in `presc.data`)
- `visit.colname`: A *string*, the name of the column in
`presc.data` containing the number of the visit or a new column name if the
prescribing data does not contain such a column.
- `split.on.dosage.change`: *Logical* or *string*. If `TRUE`
split the dispensing event on days with dosage change and create a new event with
the new dosage for the remaining supply. If *string*, the name of the column
containing the `Logical` in *disp.data* for each medication class separatly.
Important if carryover should be considered later on. See details below.
- `force.init.presc`: *Logical*. If `TRUE` advance the date of the
first prescription event to the date of the first dispensing event, if the first
prescription event is after the first dispensing event for a specific medication.
Only if the first prescription event is not limited in duration (as indicated in
the `presc.duration.colname`). See details below.
- `force.presc.renew`: *Logical* or *string*. If `TRUE` require
a new prescription for all medications for every prescription event (visit),
otherwise prescriptions end on the first visit without renewal. If *string*,
the name of the column in *disp.data* containing the `Logical` for each
medication class separatly. See details below.
- `trt.interruption`: Can be either of *"continue"*, *"discard"*,
*"carryover"*, or a *string*. It indicates how to handle durations during
treatment interruptions (see `special.periods.mapping`).
If *string*, the name of the (*character*) column in *disp.data*
containing the information (*"continue"*, *"discard"*, or *"carryover"*)
for each medication class separatly. See details below.
- `special.periods.method`: A *string*, can be either of *continue*, *discard*,
*carryover*, or *custom*. It indicates how to handle durations during special periods.
With *continue*, special periods have no effect on durations and event start dates.
With *discard*, durations are truncated at the beginning of special periods and the
remaining quantity is discarded. With *carryover*, durations are truncated
at the beginning of a special period and a new event with the remaining duration
is created after the end of the end of the special period. With *custom*, the
mapping has to be included in *special.periods.data*.
- `date.format` : A *string* of the date format (identical in all data sources)
- `suppress.warnings`: *Logical*, if `TRUE` don't show any warnings.
- `return.data.table`: *Logical*, if `TRUE` return a
`data.table` object, otherwise a `data.frame`.
- `progress.bar`: *Logical*, if `TRUE` show a progress bar.
- `...`: other possible parameters.

### Exercise: link dispensing and prescription data with `AdhereR`

Use the `compute_event_durations` function to merge `durcomp.dispensing` and `durcomp.prescribing` into `event_duration_list`. Medications should be grouped by `ATC.CODE`, `UNIT`, and `FORM` (`medication.class.colnames = c("ATC.CODE", "UNIT", "FORM")`).
We don't want to use the first prescribed dose for dispensing events occuring before the first prescription event (`force.init.presc = FALSE`), but a prescription should end if a treatment has not been prescribed during a prescription event (`force.presc.renew = TRUE`) and we want a new event in case of dosage changes (`split.on.dosage.change = TRUE`). Treatment interruptions should have no impact on supply durations an start dates (`trt.interruption = "continue"`). We also don't need a progress bar (`progress.bar = FALSE`).
Please note that the function takes some time to finish! You can use `summary(event_duration_list$event_durations)` to display a summary of the results.

```{r link-multiple-sources, exercise=TRUE}
# Merge dispensing and prescription data with AdhereR's medication_match function:

```

```{r link-multiple-sources-solution}
# Merge dispensing and prescription data with AdhereR's medication_match function:
event_duration_list <- compute_event_durations(disp.data = durcomp.dispensing, 
                                               presc.data = durcomp.prescribing, 
                                               ID.colname = "ID", 
                                               medication.class.colnames = c("ATC.CODE", "UNIT", "FORM"), 
                                               disp.date.colname = "DATE.DISP", 
                                               total.dose.colname = "TOTAL.DOSE",
                                               presc.date.colname = "DATE.PRESC", 
                                               presc.daily.dose.colname = "DAILY.DOSE", 
                                               presc.duration.colname = "PRESC.DURATION", 
                                               visit.colname = "VISIT", 
                                               force.init.presc = FALSE, 
                                               force.presc.renew = TRUE, 
                                               split.on.dosage.change = TRUE, 
                                               trt.interruption = "continue", 
                                               date.format = "%Y-%m-%d",suppress.warnings = FALSE, 
                                               progress.bar = FALSE,
                                               return.data.table = TRUE)
summary(event_duration_list$event_durations)
```

### Output of *compute_event_durations*

The output of `compute_event_durations` is a list containing all data required for CMA computations, plus additional information:

- `event_durations`: a `data.table` or `data.frame` with the event durations and all required columns for CMA computations, plus 
    + `SPECIAL.DURATION`: the number of days *during* the current duration affected by special durations or treatment interruptions of type "continue"
    + `CARRYOVER.DURATION`: the number of days *after* the current duration affected by special durations or treatment interruptions of type "carryover"
    + `tot.presc.interruptions`: the total number of prescription interruptions per patient for a specific medication.
    + `tot.dosage.changes`: the total number of dosage changes per patient for a specific medication.
- `prescription_episodes`: a `data.table` or `data.frame` with the prescription episodes
- `special_periods`: a `data.table` or `data.frame` with the special periods (if applicable) 

In addition, the output contains all the arguments to the function call: 

- `special.periods.mapping`
- `ID.colname`
- `medication.class.colnames`
- `disp.date.colname`
- `total.dose.colname`
- `presc.date.colname`
- `presc.daily.dose.colname`
- `presc.duration.colname`
- `visit.colname`
- `split.on.dosage.change`
- `force.init.presc`
- `force.presc.renew`
- `trt.interruption`
- `special.periods.method`
- `date.format`

Table 3 shows the event durations for the first patient:
```{r, include = FALSE}
event_durations_list <- compute_event_durations(disp.data = durcomp.dispensing, 
                                                presc.data = durcomp.prescribing, 
                                                ID.colname = "ID", 
                                                presc.date.colname = "DATE.PRESC", 
                                                disp.date.colname = "DATE.DISP", 
                                                date.format = "%Y-%m-%d",
                                                medication.class.colnames = c("ATC.CODE", "UNIT", "FORM"), 
                                                total.dose.colname = "TOTAL.DOSE", 
                                                presc.daily.dose.colname = "DAILY.DOSE", 
                                                presc.duration.colname = "PRESC.DURATION", 
                                                visit.colname = "VISIT", 
                                                force.init.presc = FALSE, 
                                                force.presc.renew = TRUE, 
                                                split.on.dosage.change = TRUE, 
                                                trt.interruption = "continue", 
                                                suppress.warnings = FALSE, 
                                                return.data.table = TRUE)
```

```{r}
event_durations_list$event_durations[ID==1]
```

### Further data preparation

In principle the output of compute_event_durations can be used directly for CMA computation (after removing rows with zero or missing durations). However, `AdhereR` offers additional functions `prune_event_durations` and `cover_special_periods` to deal with specific cases. Please refer to the vignette [Data preparation for computing adherence to medication in AdhereR](https://ddediu.github.io/AdhereR/compute_event_durations/compute_event_durations.html){:target="_blank"} for additional information.

## Visualization of patient records

### Explore medication histories 

A first step towards deciding which algorithm is appropriate for these data is to explore medication histories visually. 

We can do this by creating an object of type `CMA0` for the example patients, and plotting it. 
This type of plots can of course be created for a much bigger subsample of patients and saved as as a `JPEG`, `PNG`, `TIFF`, `EPS` or `PDF` file using `R`'s plotting system for data exploration.

```{r, echo=TRUE}
# select to patients to plot (7 and 16). Remove rows where DURATION is NA.
example_data <- na.omit(event_durations[ID %in% c(7,16)], cols = "DURATION")

cma0 <- CMA0(data=example_data, # use the two selected patients
             ID.colname="ID", # the name of the column containing the IDs
             event.date.colname="DISP.START", # the name of the column containing the event start date
             event.duration.colname="DURATION", # the name of the column containing the duration
             event.daily.dose.colname="DAILY.DOSE", # the name of the column containing the dosage
             medication.class.colname="ATC.CODE"); # the name of the column containing the medication class
```

```{r plot-cma0-setup}
# select to patients to plot (7 and 16). Remove rows where DURATION is NA.
example_data <- na.omit(event_durations[ID %in% c(7,16)], cols = "DURATION")

cma0 <- CMA0(data=example_data, # use the two selected patients
             ID.colname="ID", # the name of the column containing the IDs
             event.date.colname="DISP.START", # the name of the column containing the event start date
             event.duration.colname="DURATION", # the name of the column containing the duration
             event.daily.dose.colname="DAILY.DOSE", # the name of the column containing the dosage
             medication.class.colname="ATC.CODE"); # the name of the column containing the medication class
```

## Exercise: Plot medication events

Use `plot()` to visualize the medication events for `example_data`. You can set `align.all.patients = TRUE` to plot each patient relative to the earliest event date. 

```{r plot-cma0, exercise = TRUE}
# Plot the cma0-object (CMA0 shows the actual event data only):

```

```{r plot-cma0-solution}
# Plot the object (CMA0 shows the actual event data only):
plot(cma0, # the object to plot
     align.all.patients=TRUE # plot all patients relative to their earliest event date.
     )
```

### Subset to medication class

As you can see, both patients have a number of different medications. We are interested in *Selective beta-2-adrenoreceptor agonists* (ATC chemical subgroup *R03AC*). To subset the data to medications in this subgroup only, we can use another useful function `grepl()`, which searches for patterns in strings:

```{r, echo = TRUE}
#subset to medications of the ATC subgroup R03AC. The '^' at the beginning of the string tells the function that the following pattern has to be at the beginning of the string.
example_data <- example_data[grepl("^R03AC",ATC.CODE)]
```


```{r subset_data}
#subset to medications of the ATC subgroup R03AC. The '^' at the beginning of the string tells the function that the following pattern has to be at the beginning of the string.
example_data <- example_data <- na.omit(event_durations[ID %in% c(7,16) & grepl("^R03AC",ATC.CODE)], cols = "DURATION")
```


### Interactive plotting

During the exploratory phases of data analysis, it is sometimes extremely useful to be able to plot interactively various views of the data using different parameter settings.
We have implemented such interactive plotting of medication histories and (simple and iterative) CMA estimates within [`RStudio`](https://www.rstudio.com) through the `plot_interactive_cma()` function. 

This function cannot be used in this interactive tutorial, but you can use RStudio to check it out!

## Adherence estimation

### Adherence, Initiation, Implementation, and Persistence

`AdhereR` facilitates the estimation of adherence during all its phases according to the ABC taxonomy. We will look at each element in the following order:

1. Adherence
2. Initiation
3. Persistence
4. Implementation

## Adherence

`AdhereR`estimates adherence as *Continuous Medication Availability* (CMA). The *simple* CMA measures *CMA1* - *CMA9* do not differentiate between implementation and persistence. 

### Exercise: Calculate adherence

Calculate CMA7 estimates for the patient with the ID 16. The follow-up window starts 30 days after the first event and the observation window should cover 450 days, starting 180 days after the first event. Plot the results as seen in the previous section.

```{r calculate-CMA7, exercise = TRUE, exercise.setup = "subset_data"}

```

```{r calculate-CMA7-solution}
cma7 <- CMA7(data=example_data[ID == 16],
             ID.colname="ID",
             event.date.colname="DISP.START",
             event.duration.colname="DURATION",
             event.daily.dose.colname="DAILY.DOSE",
             medication.class.colname="ATC.CODE",
             carry.only.for.same.medication=TRUE,
             consider.dosage.change=TRUE,
             followup.window.start=30,
             observation.window.start=180,
             observation.window.duration=450); 

# Plot the object:
plot(cma7) # the object to plot
```

##  Initiation

If we have a dispensing and a prescription database, we can calculate time to initiation with the new `AdhereR` function `time_to_initiation`. 

### Exercise: Calculate time to initiation

Compute time to initiation with the function `time_to_initiation()` for the two patients in `example_data`. Select only medications with the ATC code *R03AC12* (salmeterol). 

```{r time-to-initiation, exercise = TRUE, exercise.setup = "subset_data"}



```

```{r time-to-initiation-solution}
time_to_initiation(presc.data = example_data[grepl("R03AC12",ATC.CODE)], 
                   disp.data = example_data[grepl("R03AC12",ATC.CODE)],
                   ID.colname = "ID", 
                   presc.start.colname = "START.PRESC", 
                   disp.date.colname = "DATE.DISP", 
                   medication.class.colnames = c("ATC.CODE", "UNIT", "FORM"))
```

## Persistence

To distinguish between persistence with treatment and quality of implementation, `AdhereR` provides the function `compute.treatment.episodes()`. It reports the patient ID, episode number, date of episode start, gap days at the end of or after the treatment episode, duration of episode, and episode end date.

### Exercise: compute treatment episodes

Use the `compute.treatment.episodes()` function to compute treatment episodes for `example_data` from the first event until the end of the follow-up period (3 years) with the following conditions:

- Change of medication type should be considered as a new treatment episode
- A gap of 90 days or more should be considered as a treatment interruption/discontinuation
- Existing supply will be finished before starting a new supply of the same medication, but not for different medications
- If the dosage recommendation changes with a new supply, existing supply will be administered according to the new dosage recommendations

```{r treatment-episodes, exercise = TRUE, exercise.eval=TRUE, exercise.setup = "subset_data"}

```

```{r treatment-episodes-solution}
TEs <- compute.treatment.episodes(example_data,
                                  ID.colname="ID",
                                  event.date.colname="DISP.START",
                                  event.duration.colname="DURATION",
                                  event.daily.dose.colname="DAILY.DOSE",
                                  medication.class.colname="ATC.CODE",
                                  carryover.within.obs.window = TRUE, # carry-over inside the OW
                                  carry.only.for.same.medication = TRUE, # & only for same type
                                  consider.dosage.change = TRUE, # dosage change starts new episode...
                                  medication.change.means.new.treatment.episode = TRUE, # & type change
                                  maximum.permissible.gap = 90, # & a gap longer than 180 days
                                  maximum.permissible.gap.unit = "days", # unit for the above (days)
                                  followup.window.start = 0, # 2-years FUW starts at earliest event
                                  followup.window.start.unit = "days",
                                  followup.window.duration = 3*365,
                                  followup.window.duration.unit = "days")

TEs
```

## Implementation

If we want to make sure to only calculate adherence during the implementation phase, AdhereR offers functions to calculate any simple CMA only during treatment episodes as computed with the `compute.treatment.episodes()` function.

### Exercise: Calculate adherence for multiple treatment episodes during an observation window.

Use the `CMA_per_episode()` function to compute CMA7 during the implementation phase for the patient with the ID 7 in `example_data` from the first event until the end of the follow-up period (3 years) with the following conditions:

- Change of medication type should be considered as a new treatment episode
- A gap of 90 days or more should be considered as a treatment interruption/discontinuation
- Existing supply will be finished before starting a new supply of the same medication, but not for different medications
- If the dosage recommendation changes with a new supply, existing supply will be administered according to the new dosage 


```{r implementation, exercise = TRUE, exercise.setup = "subset_data"}


```

```{r implementation-solution}
# Calculate CMA7:
cmaE <- CMA_per_episode(CMA="CMA7", # apply the simple CMA7 to each treatment episode
                        data=example_data[ID == 7],
                        ID.colname="ID",
                        event.date.colname="DISP.START",
                        event.duration.colname="DURATION",
                        event.daily.dose.colname="DAILY.DOSE",
                        medication.class.colname="ATC.CODE",
                        carryover.within.obs.window = TRUE,
                        carry.only.for.same.medication = TRUE,
                        consider.dosage.change = TRUE, 
                        medication.change.means.new.treatment.episode = TRUE,
                        maximum.permissible.gap = 90,
                        maximum.permissible.gap.unit = "days",
                        followup.window.start=0,
                        followup.window.start.unit = "days",
                        followup.window.duration = 365 * 3,
                        followup.window.duration.unit = "days",
                        observation.window.start=0,
                        observation.window.start.unit = "days",
                        observation.window.duration=365*3,
                        observation.window.duration.unit = "days")

plot(cmaE, show.legend=FALSE)
```




