---
title: "Importing datasets and performing descriptives"
author: "Alex Dima"
date: " `r Sys.Date()`"
output: 
    html_document: 
    highlight: textmate
    theme: flatly
    toc: TRUE
    tables: TRUE
---

# Summary 

This document imports several datasets with which we will work during the data analysis workshop. You can explore one of these datasets (or import and explore your own dataset) using what you have learned from the abccode script and the new functions presented below.


```{r setup, warning=FALSE, message=FALSE, echo=FALSE}

# install and upload some useful libraries ####

for (n in c('readxl', 
            'tableone', 'ggplot2', 
            'AdhereR',
            'dplyr', 'tidyr',
            'knitr', 'bookdown'))
{
  if(!require(n,character.only=TRUE)){install.packages(n)}
  library(n,character.only=TRUE)
}

```

Three datasets are imported, four other are available within the AdhereR package.

```{r dataset, include=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
# import data from other formats and check it 

# for SR data analysis
## from excel:
data.3Qs <- read_excel("SR_BIN1_data.xlsx", 2) # read in the 2nd worksheet from the workbook SR_BIN1_data.xlsx, first row contains variable names (default)
# check data 
# View(data.3Qs)
# OR
head(data.3Qs)


# the data for the EM data analysis
### from csv:
data.onco <- read.csv2("Onco.csv",na.string = "",header=T)  
head(data.onco)
data.RIVA <- read.csv2("RIVA.csv",na.string = "",header=T)  
head(data.RIVA)

# for the EHD analysis, datasets are available with the AdhereR package
head(med.events)
head(durcomp.prescribing)
head(durcomp.dispensing)
head(durcomp.hospitalisation)

# explore the self-report dataset
# check sample size
nrow(data.3Qs)
# check variable names
names(data.3Qs)
# check number of unique IDs (sample size)
length(unique(data.3Qs$DATSTAT_ALTPID_fix))
# check number of observations per patient
table(table(data.3Qs$DATSTAT_ALTPID_fix))
# check number of visits per patient
table(data.3Qs$visitnumber)

```

The datasets imported have the following sample sizes: 

- `r length(unique(data.onco$id))` participants (the Onco dataset), 
- `r length(unique(data.onco$id))` participants (the RIVA dataset), 
- `r length(unique(data.3Qs$DATSTAT_ALTPID_fix))` participants (the Self-report dataset).

## Descriptives

Response frequencies and descriptive statistics are examined for some variables in the Self-report dataset. You can do the same with the other datasets imported.

```{r, include=FALSE, warning=FALSE, echo=FALSE, message=FALSE}

# run frequencies for all items
for( n in names(data.3Qs))
{
  cat( "\n", n, ":" );
  print( table( data.3Qs[,n], exclude=NULL ) );
}

# check how a variable is coded
class(data.3Qs$gender)
# check the frequencies for a single variable
table(data.3Qs$gender, exclude=NULL)
# transform into factor variable
data.3Qs$genderF <- as.factor(data.3Qs$gender)
# check frequencies
table(data.3Qs$genderF, exclude=NULL)
# check frequencies when every patient is selected once (visit)
table(data.3Qs$genderF[data.3Qs$visitnumber==1], exclude=NULL)

##-------------##
#### TO DO 1 ####

# check descriptives for SDEM1_AGE
# min()
# max()
# range()
# median()
# quantile()
# mean()
# sd()

##-------------##

# demographics descriptives: table one
myVars <- c("genderF",  "SDEM1_AGE")
## Vector of categorical variables that need transformation
catVars <- c("genderF")
## Create a TableOne object
tab1 <- CreateTableOne(vars = myVars, # all variables to include in table
                       data = data.3Qs[data.3Qs$visitnumber==1,], # the dataset
                       factorVars = catVars #, # categorical variables
                      #strata = "StudyF" # which variable to use for stratifying the table
                      ) 
tabDemogr <- print(tab1, 
      showAllLevels = TRUE, # if all factor levels should be included in the table for categorical variables
      # SMD=TRUE # to show standardized mean differences
      test=FALSE)

```

The table below shows descriptive statistics for gender and age of the self-report sample:

```{r, results = 'asis', warning=FALSE, echo=FALSE, message=FALSE}
knitr::kable(tabDemogr)

```

A histogram is shown here for the age variable, with the graphics and the ggplot2 packages

```{r, fig.width=5, fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}

# plotting
hist(data.3Qs$SDEM1_AGE[data.3Qs$visitnumber==1], 
     main = "Histogram of Age")

# with ggplot
myAGEplot <-
  ggplot(data.3Qs[data.3Qs$visitnumber == 1, ], aes(x = SDEM1_AGE)) + 
  geom_histogram() 
myAGEplot

##-------------##
#### TO DO 2 ####

# change the title of the basic histogram:
# add a label for the x axis by looking at the hist function help

# change the looks of the ggplot histogram:
# change design & add axis labels: see what happens if you add to myAGEplot the following 
# + theme_classic() + labs( x = "years", y = "number of participants")

# look at the data viz cheatsheet and add colors to myAGEplot

##-------------##

```



