---
title: "Descriptive statistics one dataset for example"
author: "Alex Dima"
date: " `r Sys.Date()`"
output: 
    html_document: 
    highlight: textmate
    theme: flatly
    toc: TRUE
    tables: TRUE
---

#Summary 

This document imports several datasets from different formats. .... more text to remember what I do

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
# this is section of R script imports and uploads some libraries which will be used later on
# here you may want to define some useful settings and functions if your analysis is more complex

# install and upload some useful libraries ####

for (n in c('foreign', 'readxl', 'utils', 'relimp', 'tableone',
            'dplyr', 'tidyr', 'reshape2', 'tibble', 'ggplot2', 
            'knitr', 'bookdown'))
{
  if(!require(n,character.only=TRUE)){install.packages(n)}
  library(n,character.only=TRUE)
}

#  brief info on the uses of these libraries:
#  for importing data: 'readr', 'readxl'
#  for data viewing, writing: 'utils', 'relimp'
#  for data management: 'dplyr', 'tidyr', 'reshape2', 'tibble', 
#  for reporting documents: 'knitr', 'bookdown'
#  for plotting: 'ggplot'

```

```{r dataset, warning=FALSE, message=FALSE, echo=FALSE}
# this is a section of R script for importing data from other formats and checking it to make sure you imported the right one
# you can modify depending on the format and name of your dataset

### import data files with an .RData format (e.g. exported at the end of data preparation in a prior R script) ####
load("my1stEnvironment.RData")

# Here are examples of how you can import data from excel, csv, and SPSS

### from excel:
# this is a dataset available on https://zenodo.org/record/854656#.Wfy4K1ynGog
campus <- read_xlsx("./DatabaseHealthCampusStudies.xlsx", sheet = 1) # read in the first worksheet from the workbook myexcel.xlsx, first row contains variable names (default)

# check data 
# View(campus)
# OR
# head(campus)
# check sample size
nrow(campus)
# check variable names
names(campus)

### from csv:
HoNOS <- read.table(file="./HoNOSData.csv",
                     header=TRUE, # first row contains variable names
                     sep=","   # comma is separator
                     ) 

# from SPSS:
# this is a dataset available on https://zenodo.org/record/16272#.Wfy47FynGog
surgery <- foreign::read.spss("./Andersen_1_2015.sav", to.data.frame = TRUE, 
  use.value.labels = FALSE, use.missings = FALSE)

# select a part of a dataset you want to work with
# for example, if the item set you want to study is the questionnaire in the HoNOS data, the code would be:
HONOSqre <- HoNOS[7:18]

##-------------##
#### TO DO 1 ####

# add the sample sizes of the other 2 datasets in the text below this R section

##-------------##

```

These datasets have the following sample sizes: `r nrow(campus)` participants (the Campus dataset), `r nrow(HoNOS)` participants (the HoNOS dataset), `r nrow(surgery)` participants (the Surgery dataset).

# Descriptives

Response frequencies and descriptive statistics are examined for a range of questions in the Campus dataset. You can do the same with the other datasets imported.


```{r, include=FALSE, warning=FALSE, echo=FALSE, message=FALSE}

# run frequencies for all items
for( n in names(campus))
{
  cat( "\n", n, ":" );
  print( table( campus[,n], exclude=NULL ) );
}

# check how a variable is coded
class(campus$Study)
# check the frequencies for a single variable
table(campus$Study, exclude=NULL)
# transform into factor variable
campus$StudyF <- as.factor(campus$Study)
# check mode
which.max(table(campus$StudyF))


# check how this variable is coded
class(campus[,"Gender 1:H 2:F"])
# change name variable
names(campus)[names(campus) == "Gender 1:H 2:F"] <- "Gender"
# check coding again
class(campus$Gender)
# check the frequencies for a single variable
table(campus$Gender, exclude=NULL)

# transform into factor variable
campus$GenderF <- as.factor(campus$Gender)
# check the frequencies again
table(campus$GenderF, exclude=NULL)

# define levels
levels(campus$GenderF) <- c("H", "F")
# check the frequencies again
table(campus$GenderF, exclude=NULL)

##-------------##
#### TO DO 2 ####

# change the name of the variable "3  BMI"  into "BMI"
names(campus)[names(campus) == "3  BMI"] <- "BMI"

# check what class is "BMI" and do a frequency table

##-------------##

# check how a variable is coded
class(campus$AGE)

##-------------##
#### TO DO 3 ####

# check descriptives for AGE
# min()
# max()
# range()
# median()
# quantile()
# mean()
# sd()

##-------------##


#table one

myVars <- c("GenderF", "BMI", "AGE")
## Vector of categorical variables that need transformation
catVars <- c("GenderF", "BMI")
## Create a TableOne object
tab1 <- CreateTableOne(vars = myVars, # all variables to include in table
                       data = campus, # the dataset
                       factorVars = catVars, # categorical variables
                      strata = "StudyF" # which variable to use for stratifying the table
                      ) 
tabQoF <- print(tab1, 
      showAllLevels = TRUE, # if all factor levels should be included in the table for categorical variables
      # SMD=TRUE # to show standardized mean differences
      test=FALSE)

rownames(tabQoF) <- c("n", "Gender (%)", "",   
                           "BMI (%)", "", "", 
                            "Age (mean (SD))"  )
tabQoF  <- cbind("Values" = rownames(tabQoF), tabQoF)
rownames(tabQoF) <- NULL
colnames(tabQoF) <- c("Indicator", "Levels", "Study1", "Study2" )
```

```{r, results = 'asis', warning=FALSE, echo=FALSE, message=FALSE}

knitr::kable(tabQoF)

```

```{r, fig.width=5, fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}

# plotting

# "Filiere 4 classes" 
# change name variable
names(campus)[names(campus) == "Filiere 4 classes"] <- "Filiere4"
pie(table(campus$Filiere4, exclude=NULL), main="Piechart Filiere 4 categories")
barplot(table(campus$Filiere4))

# pour les continues
boxplot(campus$AGE)
hist(campus$AGE)

# avec ggplot
myAGEplot <- ggplot(campus, aes(x=AGE)) + geom_histogram()
myAGEplot

##-------------##
#### TO DO 4 ####


# do 3 barplots for 3 variables 
# find the variable and check format
# do any necessary transformations / labelling
# do the barplots with basic functions
# do the barplots with ggplot

# change the size of the plots

# add titles, subtitles, labels, colors to all plots


##-------------##


```



